Comparison of long and short amplicon metabarcoding for determination of fungal species diversity and spatial turnover

### 2019-08-21 writing/MSA2019_presentation.Rnw
final changes before MSA presentation

\title{Comparison of long and short amplicon metabarcoding for determination of fungal species diversity and spatial turnover}
\subtitle{in an ectomycorrhizal West African woodland}
\author{\mbox{\textbf{Brendan Furneaux}\inst{1}} \and \mbox{Mohammad Bahram\inst{2}} \and \mbox{Nourou Yorou\inst{3}} \and \mbox{Anna Rosling\inst{1}} \and \mbox{Martin Ryberg\inst{1}}}
\institute{\inst{1}Uppsala University \and \inst{2}Swedish University of Agricultural Sciences \and \inst{3}University of Parakou}
\date{13 August, 2019}
\logo{\includegraphics[width=2.5cm]{Uppsala_University_logo}}

@@

<<setup, include=FALSE>>=
knitr::opts_chunk$set(echo = FALSE, device = "tikz", cache = TRUE)
library(drake)
library(magrittr)
library(tidyverse)
library(ggplot2)
library(scales)
library(ggtree)
library(castor)
theme_set(theme_bw(base_size = 24) + theme(legend.direction = "vertical"))
loadd(bc_corr, bc_corr2, unif_corr, unif_corr2, qstats_join, datasets, longtree,
      taxon_LSUcons_rdp)
reads <- qstats_join %>%
  mutate_at("Region", factor,
            levels = c("ITS2", "ITS1", "short", "ITS", "LSU", "long")) %>%
  mutate_at("Dataset", factor,
            levels = c("short-ion", "short-pacbio", "long-pacbio")) %>%
  group_by(Dataset, Plate, Region)
reads_subset <- reads %>%
  do(if (nrow(.) > 10000) sample_n(., 10000) else .)
reverselog_trans <- function(base = exp(1)) {
    trans <- function(x) -log(x, base)
    inv <- function(x) base^(-x)
    trans_new(paste0("reverselog-", format(base)), trans, inv, 
              log_breaks(base = base), 
              domain = c(1e-100, Inf))
}
@

\setbeamersize{text margin left=5mm,text margin right=5mm}

\begin{frame}{rDNA Regions and Databases}
\framesubtitle{rDNA regions}
\vskip -1ex

@@

\vfill

@@

\onslide<2>
\begin{columns}[t]
  \begincol{0.3\linewidth}
    \begin{block}{Silva 132 (2017)}
      \small
      \textbf{SSU Ref:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 2M seqs
        \item 700k "Ref NR 99" (clusters)
        \item 16k Fungi, 60k other Eukaryota, 625k Prokaryota
      \end{itemize}
      \textbf{LSU Ref:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 200k seqs
        \item 5k Fungi, 24k other Eukaryota, 170k Prokaryota
      \end{itemize}
    \end{block}
  \endcol
  \begincol{0.3\linewidth}
    \begin{block}{Unite 8.0 (2019)}
    \small
    \textbf{ITS:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 800k seqs
        \item 120k "species hypotheses" (clusters)
        \item 70k Fungi, 40k plants, 10k other Eukaryota
        \itemsome sequences include LSU
      \end{itemize}
    \end{block}
  \endcol
  \begincol{0.3\linewidth}
    \small
    \begin{block}{RDP 11 (2016)}
      \small
      \textbf{Fungal LSU training set:}\\
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 11k seqs
        \item 8k Fungi, 3k other Eukaryota
        \item annotated to genus
      \end{itemize}
      \textbf{SSU dataset:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item Prokaryotes only
      \end{itemize}
      \textbf{Warcup Fungal ITS training set:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 18k seqs
        \item all Fungi
      \end{itemize}
    \end{block}
  \endcol
\end{columns}
\center The three databases don't use the same taxonomy!
\end{overprint}
\end{frame}

@@

\begin{frame}{Introduction}

\begin{itemize}
 \item Question: What is the scale of spatial (and temporal) autocorrelation in soil fungal community composition?
  \begin{itemize}
    \item Specifically in a tropical ECM forest.
  \end{itemize}
 \item Metabarcoding method comparison
\end{itemize}
\end{frame}

\begin{frame}{Study site}
\vskip1.5em
\begincols
  \begincol{0.55\linewidth}
    \begin{figure}
      \begin{overprint}
      \onslide<1>
        \includegraphics{soudanian_benin}
      \onslide<2>
        \includegraphics{soudanian_benin_oueme}
      \end{overprint}
      \caption{Location of study sites in Bènin.}
    \end{figure}
  \endcol

@@

\begin{frame}{Sampling}
  \begin{itemize}
    \item<1-|alert@1> \SI{25}{\m} transects.
    \item<1-|alert@1> Soil samples taken every 1 m.
    \item<2-|alert@2> $\SI{5}{\cm}\times\SI{5}{\cm}\times\SI{5}{\cm}$ sample, homogenize in plastic bag, then subsample \SI{50}{\mg}
    \item<3-|alert@3> Lysis \& preservation in field
    \item<4-|alert@4> Same locations in two years -- 2015 and 2016 (25 points × 2 years × 2 sites = 100 samples)
  \end{itemize}
\end{frame}

\begin{frame}{Sequencing}
  \framesubtitle{Which region?}
  \small
  \includegraphics[width=\linewidth]{rDNA_amplicons}

  \begincols
  \begincol{0.5\linewidth}
    \begin{block}{ITS2 (gITS7-ITS4)}
        \begin{itemize}
          \tightlist
          \item Internally transcribed spacer between 5.8S and 26/28S (LSU) subunits of ribosomal RNA
          \item Usually 250--350 bp.
          \item ITS --- standard DNA barcode for fungi.
          \item Relatively good species-level resolution, and good database support.
          \item Cannot be aligned across larger taxonomic groups.
          \item Groups with variant (or absent) 5.8S not sequenced.
        \end{itemize}
      \end{block}
    \endcol
    \begincol{0.5\linewidth}
      \begin{block}{Full ITS + partial LSU (ITS1-LR5)}
        \begin{itemize}
          \tightlist
          \item Full ITS plus the 5' end of LSU; 1000-2000 bp.
          \item ITS -- resolution to separate closely related taxa.
          \item LSU -- can be aligned across Fungi.
          \item Primers are more universal than gITS7.
          \item Long amplicon with alternating variable and conserved regions --- susceptible to chimera formation and breakage.
        \end{itemize}
      \end{block}
    \endcol
  \endcols
\end{frame}

% \begin{frame}{Sequencing}
%   \framesubtitle{Which platform?}
%   \begincols
%   \begincol{0.5\linewidth}
%     \begin{block}{IonTorrent}
%       \begin{itemize}
%         \tightlist
%         \item Low cost per bp.
%         \item Smaller minimum job than Illumina.
%         \item Needs less DNA than PacBio.
%         \item Bias against amplicons greater than 400 bp.
%          \item Cannot sequence long ITS+LSU amplicon.
%        \end{itemize}
%      \end{block}
%   \endcol
%   \begincol{0.5\linewidth}
%      \begin{block}{PacBio}
%         \begin{itemize}
%           \tightlist
%           \item Higher cost per bp $\implies$ fewer samples or less sequencing depth for same budget.
%           \item Requires more DNA.
%           \item Good accuracy requires multiple circular reads.
%         \end{itemize}
%       \end{block}
%   \endcol
%   \endcols
%   \centering
% 
% \end{frame}
% 
% \begin{frame}{Sequencing}
%   \begin{block}{The plan}
%   Three sequencing runs:
%   \begin{itemize}
%     \item Ion Proton S5, short amplicon
%     \begin{itemize}
%       \item Pool all samples on one chip
%     \end{itemize}
%     \item PacBio RSII, short amplicon
%     \begin{itemize}
%       \item Same amplicon library as IonTorrent.
%       \item Two sequencing libraries, two SMRT cells per library.
%     \end{itemize}
%     \item PacBio RSII, long amplicon
%     \begin{itemize}
%       \item Two sequencing libraries, four SMRT cells per library.
%     \end{itemize}
%   \end{itemize}
%   \end{block}
% \end{frame}
% 
% \begin{frame}{Sequencing}
%   \begin{block}{What happened}
%   Three sequencing runs:
%   \begin{itemize}
%     \item \sout{IonTorrent, short amplicon}
%     \begin{itemize}
%       \item Accidentally used primers with IonTorrent adapters built in.
%       \item Lost some of the multiplexing barcodes.
%       \item Can't demultiplex the samples --- can only look at pool of all reads.
%     \end{itemize}
%     \item PacBio, short amplicon
%     \begin{itemize}
%       \item OK
%     \end{itemize}
%     \item PacBio, long amplicon
%     \begin{itemize}
%       \item OK
%     \end{itemize}
%   \end{itemize}
%   \end{block}
% \end{frame}

\begin{frame}{Bioinformatics}
  \begin{itemize}
    \item Redo RSII circular consensus basecalling using Sequel algorithm (Arrow).
    \item Demultiplex and trim primers.
    \item Cut out regions (ITS1, ITS2, full ITS, LSU) using ITSx.
    \item Quality filter - up to 3 expected errors.
  \end{itemize}
\end{frame}

\begin{frame}{Expected error rate}
\framesubtitle{After \texttt{Arrow}}

<<errorrate, fig.width = 12, cache.check = reads_subset>>=
filter(reads_subset, Tech == "PacBio") %>%
  ggplot(aes(x = eexp,
             group = paste(Dataset, Region),
             color = Dataset)) +
  geom_line(aes(y = 1 - ..y..), stat = "ecdf", size = 1) +
  scale_x_continuous(name = "Expected number of errors",
                     trans = reverselog_trans(10),
                     limits = c(NA, 1e-5),
                     oob = squish) +
  facet_wrap(~Region, ncol = 2, strip.position = "right") +
  scale_y_continuous(name = "Fraction passing") +
  # scale_color_viridis_d(option = "C") +
  geom_vline(aes(xintercept = 3), linetype = 2)
@
\end{frame}

\begin{frame}{Bioinformatics}
\framesubtitle{continued...}
  \begin{itemize}
    \item Find amplicon sequence variants (ASVs) with \texttt{dada2}.
      \begin{itemize}
        \item Doesn't work well on full long amplicons -- algorithm depends on having a large number of error-free reads.
        \item Works on the smaller regions.
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Expected total number of errors}
\framesubtitle{After \texttt{Arrow}}

<<length, fig.width = 12, cache.check = reads_subset>>=
filter(reads_subset, Tech == "PacBio", eexp <= 3) %>%
  ggplot(aes(x = length,
             group = paste(Dataset, Region),
             color = Dataset, fill = Dataset)) +
  geom_area(stat = "density", size = 1, alpha = 0.5) +
  scale_x_log10(name = "Length (bp)",
                limits = c(50, NA),
                oob = squish) +
  facet_wrap(~Region, ncol = 2, strip.position = "right", scales = "free_y") +
  scale_y_continuous(name = "Fraction passing") +
  # scale_color_viridis_d(option = "C") +
  geom_vline(aes(xintercept = 3), linetype = 2)
@
\end{frame}

\begin{frame}{Bioinformatics}
\framesubtitle{continued...}
  \begin{itemize}
    \item Find amplicon sequence variants (ASVs) with \texttt{dada2}.
      \begin{itemize}
        \item Doesn't work well on full long amplicons -- algorithm depends on having a large number of error-free reads.
        \item Works on the smaller regions.
      \end{itemize}
    \item For each ITS2 ASV, align the whole long read and find the consensus.
      \begin{itemize}
        \item There were a lot of chimeras (15--25\%).
        \item Remove "outliers" from the alignments using \texttt{odseq}.
        \item Remaining chimeras are too few to change the consensus.
      \end{itemize}
    \item Align with \texttt{DECIPHER}
      \begin{itemize}
        \item Progressive aligner with inferred RNA secondary structure.
      \end{itemize}
    \item \texttt{RAxML} tree
  \end{itemize}
\end{frame}

\begin{frame}{Taxonomy assignment}
Both \texttt{dada2} and \texttt{DECIPHER} can do taxonomy assignment.
\begincols
\begincol{0.5\framewidth}
\begin{block}{dada2}
  \begin{itemize}
    \item Naïve Bayesian classifier.
    \item Uses a lot of memory.
    \item Needs every rank to be filled (\emph{incertae sedis}, "unidentified").
    \item Seems to put thins in places that don't match the tree.
  \end{itemize}
\end{block}
\endcol

\begincol{0.5\framewidth}
\begin{block}{DECIPHER}
  \begin{itemize}
    \item Some kind of machine learning algorithm.
    \item Uses less memory.
    \item Allows uneven classification depths.
    \item Very low bootstraps, especially for ITS.
  \end{itemize}
\end{block}
\endcol
\endcols

UNITE for ITS, RDP for LSU...  But there is no large database that has both for the same specimens!

I'm still working on it...
\end{frame}

\begin{frame}{Tree}
\framesubtitle{Kingdoms}
<<kingtree, fig.width = 12, message=FALSE>>=
seqhash <- function(seq, len = 8) UseMethod("seqhash")
seqhash.character <- function(seq, len = 8) {
  h <- purrr::map_chr(seq, digest::digest)
  stringr::str_sub(h, end = len)
}
seqhash.XStringSet <- function(seq, len = 8) {
  seqhash.character(as.character(seq))
}
treetax <- taxon_LSUcons_rdp %>%
  mutate_at("seq", seqhash) %>%
  select(-Taxonomy, -nreads) 
outgroups <- treetax %>% filter(!is.na(Kingdom), Kingdom %in% c("Alveolata", "Chromista", "Plantae"), seq %in% longtree$tip.label)
p <- ggtree(tr = longtree)
longtree <- reroot(longtree, MRCA(p, outgroups$seq))

ggtree(groupOTU(longtree, split(treetax$seq, treetax$Kingdom))) +
  aes(color = group) + scale_color_brewer(type = "qual", palette = "Paired") +
  theme(legend.position = NULL)
@
\end{frame}

\begin{frame}{Tree}
\framesubtitle{Phyla}
<<phytree, fig.width = 12, message=FALSE>>=
ggtree(groupOTU(longtree, split(treetax$seq, treetax$Phylum))) +
  aes(color = group) + scale_color_brewer(type = "qual", palette = "Paired") +
  theme(legend.position = NULL)
@
\end{frame}

\begin{frame}{Tree}
\framesubtitle{Families --- Basidiomycota}
<<basidtree, fig.width = 12, message=FALSE>>=
basids <- treetax %>% filter(!is.na(Class),
                             Class %in% c("Geminibasidiomycetes",
                                         "Tremellomycetes",
                                         "Exobasidiomycetes",
                                         "Microbotryomycetes"),
                             seq %in% longtree$tip.label)

basidtree <- phylobase::phylo4(longtree) %>%
  phylobase::subset(mrca = basids$seq) %>%
  as("phylo")

basids2 <- treetax %>% filter(seq %in% basidtree$tip.label)

ggtree(groupOTU(basidtree, split(basids2$seq, basids2$Order)), layout = "circular") +
  aes(color = group) +
  scale_color_brewer(type = "qual", palette = "Paired", limits = unique(na.omit(basids2$Order))) +
  theme(legend.position = NULL)
@
\end{frame}

\begin{frame}{Correlogram}
<<correlogram, fig.width = 12>>=
bind_rows(as.data.frame(bc_corr$mantel.res) %>%
            mutate(tdist = "same", metric = "bray"),
          as.data.frame(bc_corr2$mantel.res) %>%
            mutate(tdist = "diff", metric = "bray"),
          as.data.frame(unif_corr$mantel.res) %>%
            mutate(tdist = "same", metric = "unifrac"),
          as.data.frame(unif_corr2$mantel.res) %>%
            mutate(tdist = "diff", metric = "unifrac")) %>%
  mutate(Significant = `Pr(corrected)` < 0.05) %>%
  ggplot(aes(class.index, Mantel.cor, group = tdist, color = tdist, shape = Significant)) +
  geom_point(size = 3) +
  geom_line(alpha = 0.3, size = 1.5) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  # geom_line(aes(y = exp(predict(bc_fit)))) +
  scale_y_continuous(limits = c(-0.05, NA), name = "Spatial Distance correlation") +
  scale_color_discrete(name = "Year") +
  xlab("distance (m)") +
  facet_wrap("metric") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1))
@
\end{frame}

>>>>

\title{Comparison of long and short amplicon metabarcoding for determination of fungal species diversity and spatial turnover}
\subtitle{in an ectomycorrhizal West African woodland}
\author{\mbox{\textbf{Brendan Furneaux}\inst{1}} \and \mbox{Mohammad Bahram\inst{2}} \and \mbox{Nourou Yorou\inst{3}} \and \mbox{Anna Rosling\inst{1}} \and \mbox{Martin Ryberg\inst{1}}}
\institute{\inst{1}Uppsala University \and \inst{2}Swedish University of Agricultural Sciences \and \inst{3}University of Parakou}
\date{13 August, 2019}
\logo{\includegraphics[width=2.5cm]{Uppsala_University_logo}}
\setbeamersize{text margin left=5mm,text margin right=5mm}

@@

<<setup, include=FALSE>>=
knitr::opts_chunk$set(echo = FALSE, device = "tikz", cache = TRUE)
library(drake)
library(magrittr)
library(tidyverse)
library(ggplot2)
library(scales)
library(ggtree)
library(castor)
theme_set(theme_bw(base_size = 24) + theme(legend.direction = "vertical"))
loadd(
  bc_corr_long,
  bc_corr_long2,
  bc_corr_short,
  bc_corr_short2,
  # unif_corr,
  # unif_corr2,
  qstats_join,
  datasets,
  # longtree#,
  # taxon_LSUcons_rdp
  )
reads <- qstats_join %>%
  mutate_at("region", factor,
            levels = c("ITS2", "ITS1", "short", "ITS", "LSU", "long")) %>%
  mutate_at("dataset", factor,
            levels = c("short-ion", "short-pacbio", "long-pacbio")) %>%
  group_by(dataset, plate, region)
reads_subset <- reads %>%
  do(if (nrow(.) > 10000) sample_n(., 10000) else .)
reverselog_trans <- function(base = exp(1)) {
    trans <- function(x) -log(x, base)
    inv <- function(x) base^(-x)
    trans_new(paste0("reverselog-", format(base)), trans, inv, 
              log_breaks(base = base), 
              domain = c(1e-100, Inf))
}
@

\begin{frame}{Outline}
\Large
\begin{itemize}
  \item Long amplicons for metabarcoding
  \item Study site \& design
  \item (early) Results
\end{itemize}
\end{frame}

\begin{frame}{rDNA Regions}
\vskip -1ex

@@

\vskip 2ex

@@

\onslide<2>
\begin{columns}[t]
  \begincol{0.3\linewidth}
    \begin{block}{Silva 132 (2017)}
      \small
      \textbf{SSU Ref:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 2M seqs
        \item 700k "Ref NR 99" (clusters)
        \item 16k Fungi, 60k other Eukaryota, 625k Prokaryota
      \end{itemize}
      \textbf{LSU Ref:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 200k seqs
        \item 5k Fungi, 24k other Eukaryota, 170k Prokaryota
      \end{itemize}
    \end{block}
  \endcol
  \begincol{0.3\linewidth}
    \begin{block}{Unite 8.0 (2019)}
    \small
    \textbf{ITS:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 800k seqs
        \item 120k "species hypotheses" (clusters)
        \item 70k Fungi, 40k plants, 10k other Eukaryota
        \item some sequences include LSU
      \end{itemize}
    \end{block}
  \endcol
  \begincol{0.3\linewidth}
    \small
    \begin{block}{RDP 11 (2016)}
      \small
      \textbf{LSU training set:}\\
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 11k seqs
        \item 8k Fungi, 3k other Eukaryota
        \item annotated to genus
      \end{itemize}
      \textbf{SSU dataset:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item Prokaryotes only
      \end{itemize}
      \textbf{Warcup Fungal ITS training set:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 18k seqs, all Fungi
      \end{itemize}
    \end{block}
  \endcol
\end{columns}
\center The three databases don't use the same taxonomy!
\end{overprint}
\end{frame}

@@

\begin{frame}{Why long amplicons?}

\begin{itemize}
 \item Combine ITS (as barcode) with LSU/SSU (phylogenetic inference)
 \item Long read technologies are more tolerant of length variation
 \begin{itemize}
    \item PCR length bias still an issue!
 \end{itemize}
 \item Very conserved primer sites reduce primer bias.
 \item (Tedersoo et al., 2018)
\end{itemize}
\end{frame}

\begin{frame}{Why not long amplicons?}
\begin{itemize}
  \item Read quality
    \begin{itemize}
      \item PacBio circular consensus gives good quality up to a few kb
      \item Oxford Nanopore is still problematic
    \end{itemize}
  \item Chimera formation in PCR.
    \begin{itemize}
      \item Alternation of conserved and variable regions is a recipe for chimera formation.
    \end{itemize}
  \item Cost -- for the same budget, Illumina gives many more samples and/or greater depth.
  \item (Kennedy et al. 2018)

\end{itemize}
\end{frame}

\begin{frame}{Does it matter?}
\begin{itemize}
  \item Do a simple ecological study using different methods, see if the results change.
  \item Measuring spatial turnover in soil fungal community.
\end{itemize}
\end{frame}

\begin{frame}{Study site}
\vskip 1.5em
\begincols
  \begincol{0.55\linewidth}
    \begin{figure}
      \begin{overprint}
      \onslide<1>
        \includegraphics{soudanian_benin}
      \onslide<2>
        \includegraphics{soudanian_benin_oueme}
      \end{overprint}
      \caption{Location of study sites in Bènin.}
    \end{figure}
  \endcol

@@

\begin{frame}{Sampling}
  \begin{itemize}
    \item \SI{25}{\m} transects.
    \item Soil samples taken every 1 m.
    \item $\SI{5}{\cm}\times\SI{5}{\cm}\times\SI{5}{\cm}$ sample, homogenize in plastic bag, then subsample \SI{50}{\mg}
    \item Lysis \& preservation in field
    \item Same locations in two years -- 2015 and 2016 (25 points × 2 years × 2 sites = 100 samples) 
  \end{itemize}
\end{frame}

\begin{frame}{Sequencing}
  \framesubtitle{Which amplicon?}
  \includegraphics[width=\linewidth]{rDNA_amplicons}
  \vfill
  \begin{columns}[t]
  \begincol{0.5\linewidth}
    \begin{block}{short amplicon}
      \begin{itemize}
        \tightlist
        \item ITS2
        \item primers gITS7 -- ITS4
        \item 250--350 bp
        \item Sequence same amplicon library with Illumina MiSeq and PacBio RSII
        \item<2-> Don't have MiSeq data yet!
       \end{itemize}
     \end{block}
  \endcol
  \begincol{0.5\linewidth}
     \begin{block}{long amplicon}
        \begin{itemize}
          \tightlist
          \item Full ITS + partial LSU
          \item Primers ITS1 -- LR5
          \item 1200 -- 1800 bp
          \item PacBio RSII
        \end{itemize}
      \end{block}
  \endcol
\end{columns}
\end{frame}
% 
% \begin{frame}{Sequencing}
%   \begin{block}{The plan}
%   Three sequencing runs:
%   \begin{itemize}
%     \item Ion Proton S5, short amplicon
%     \begin{itemize}
%       \item Pool all samples on one chip
%     \end{itemize}
%     \item PacBio RSII, short amplicon
%     \begin{itemize}
%       \item Same amplicon library as IonTorrent.
%       \item Two sequencing libraries, two SMRT cells per library.
%     \end{itemize}
%     \item PacBio RSII, long amplicon
%     \begin{itemize}
%       \item Two sequencing libraries, four SMRT cells per library.
%     \end{itemize}
%   \end{itemize}
%   \end{block}
% \end{frame}
% 
% \begin{frame}{Sequencing}
%   \begin{block}{What happened}
%   Three sequencing runs:
%   \begin{itemize}
%     \item \sout{IonTorrent, short amplicon}
%     \begin{itemize}
%       \item Accidentally used primers with IonTorrent adapters built in.
%       \item Lost some of the multiplexing barcodes.
%       \item Can't demultiplex the samples --- can only look at pool of all reads.
%     \end{itemize}
%     \item PacBio, short amplicon
%     \begin{itemize}
%       \item OK
%     \end{itemize}
%     \item PacBio, long amplicon
%     \begin{itemize}
%       \item OK
%     \end{itemize}
%   \end{itemize}
%   \end{block}
% \end{frame}

\begin{frame}{Bioinformatics}
  \begin{itemize}
    \item Redo RSII circular consensus basecalling using Sequel algorithm (Arrow).
    \item Cut out regions (ITS1, ITS2, full ITS, LSU) using ITSx.
    \item Quality filter - up to 3 expected errors.
  \end{itemize}
\end{frame}

\begin{frame}{Expected error rate}
\framesubtitle{After \texttt{Arrow} and \texttt{ITSx}}

<<errorrate, fig.width = 12, cache.check = reads_subset>>=
filter(reads_subset, tech == "PacBio") %>%
  ggplot(aes(x = eexp,
             group = paste(dataset, region),
             color = dataset)) +
  geom_line(aes(y = 1 - ..y..), stat = "ecdf", size = 1) +
  scale_x_continuous(name = "Expected number of errors",
                     trans = reverselog_trans(10),
                     limits = c(NA, 1e-5),
                     oob = squish) +
  facet_wrap(~region, ncol = 2, strip.position = "right") +
  scale_y_continuous(name = "Fraction passing") +
  # scale_color_viridis_d(option = "C") +
  geom_vline(aes(xintercept = 3), linetype = 2)
@
\end{frame}

\begin{frame}{Bioinformatics}
\framesubtitle{continued...}
  \begin{itemize}
    \item Find ITS2 amplicon sequence variants (ASVs) with \texttt{dada2}.
    \item Generate LSU consensus for each ASV
  \end{itemize}
\end{frame}

\begin{frame}{Amplicon and individual region lengths}
\framesubtitle{After \texttt{Arrow} and \texttt{ITSx}}

<<length, fig.width = 12, cache.check = reads_subset>>=
filter(reads_subset, tech == "PacBio", eexp <= 3) %>%
  ggplot(aes(x = length,
             group = paste(dataset, region),
             color = dataset, fill = dataset)) +
  geom_area(stat = "density", size = 1, alpha = 0.5) +
  scale_x_log10(name = "Length (bp)",
                limits = c(50, NA),
                oob = squish) +
  facet_wrap(~region, ncol = 2, strip.position = "right", scales = "free_y") +
  scale_y_continuous(name = "Fraction passing") +
  # scale_color_viridis_d(option = "C") +
  geom_vline(aes(xintercept = 3), linetype = 2)
@
\end{frame}

\begin{frame}{Long vs. short amplicons}
\framesubtitle{ASV read counts}
\begin{overprint}
\onslide<1>
Total ASVs:
\begin{itemize}
  \item Long amplicon: 646 ASVs
  \item Short amplicon: 531 ASVs
\end{itemize}

\onslide<2>
<<readplot, fig.width = 12, message=FALSE>>=
loadd(big_seq_table_ITS2)
asv_table <- big_seq_table_ITS2 %>%
  tibble::as_tibble(rownames = "filename") %>%
  tidyr::gather(key = "seq", value = "reads", -1) %>%
  dplyr::filter(reads >= 1) %>%
  tidyr::extract(col = "filename",
                 into = c("seq_run", "plate", "well", "dir", "region"),
                 regex = "([a-z]+_\\d+)_(\\d+)-([A-H]1?[0-9])([fr]?)-([:alnum:]+).+") %>%
  dplyr::filter(seq_run %in% c("pb_483", "pb_500")) %>%
  dplyr::group_by(seq_run, seq) %>%
  dplyr::summarize(reads = sum(reads)) %>%
  tidyr::spread(key = seq_run, value = reads, fill = 0) %>%
  dplyr::rename(short_pacbio = pb_483, long_pacbio = pb_500)

asv_table %>% 
  filter(short_pacbio > 0, long_pacbio > 0) %>%
  ggplot(aes(x = short_pacbio, y = long_pacbio)) +
  geom_point(alpha = 0.2) +
  scale_x_log10() +
  scale_y_log10() +
  xlab("Short amplicon reads") +
  ylab("Long amplicon reads") +
  geom_rug(aes(y = long_pacbio, x = 1), sides = "l", alpha = 0.2,
           data = asv_table %>% filter(short_pacbio == 0)) +
  geom_rug(aes(y = short_pacbio), sides = "b", alpha = 0.2,
           data = asv_table %>% filter(long_pacbio == 0)) +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_abline(aes(intercept = mean(log10(long_pacbio[long_pacbio >= 1])) -
                    mean(log10(short_pacbio[short_pacbio >= 1])), slope = 1),
              linetype = 2L, alpha = 0.5)
@
\end{overprint}
\end{frame}

% \begin{frame}{Tree}
% \framesubtitle{Kingdoms}
% <<kingtree, fig.width = 12, message=FALSE>>=
% seqhash <- function(seq, len = 8) UseMethod("seqhash")
% seqhash.character <- function(seq, len = 8) {
%   h <- purrr::map_chr(seq, digest::digest)
%   stringr::str_sub(h, end = len)
% }
% seqhash.XStringSet <- function(seq, len = 8) {
%   seqhash.character(as.character(seq))
% }
% treetax <- taxon_LSUcons_rdp %>%
%   mutate_at("seq", seqhash) %>%
%   select(-Taxonomy, -nreads) 
% outgroups <- treetax %>% filter(!is.na(Kingdom), Kingdom %in% c("Alveolata", "Chromista", "Plantae"), seq %in% longtree$tip.label)
% p <- ggtree(tr = longtree)
% longtree <- reroot(longtree, MRCA(p, outgroups$seq))
% 
% ggtree(groupOTU(longtree, split(treetax$seq, treetax$Kingdom))) +
%   aes(color = group) + scale_color_brewer(type = "qual", palette = "Paired") +
%   theme(legend.position = NULL)
% @
% \end{frame}
% 
% \begin{frame}{Tree}
% \framesubtitle{Phyla}
% <<phytree, fig.width = 12, message=FALSE>>=
% ggtree(groupOTU(longtree, split(treetax$seq, treetax$Phylum))) +
%   aes(color = group) + scale_color_brewer(type = "qual", palette = "Paired") +
%   theme(legend.position = NULL)
% @
% \end{frame}
% 
% \begin{frame}{Tree}
% \framesubtitle{Families --- Basidiomycota}
% <<basidtree, fig.width = 12, message=FALSE>>=
% basids <- treetax %>% filter(!is.na(Class),
%                              Class %in% c("Geminibasidiomycetes",
%                                          "Tremellomycetes",
%                                          "Exobasidiomycetes",
%                                          "Microbotryomycetes"),
%                              seq %in% longtree$tip.label)
% 
% basidtree <- phylobase::phylo4(longtree) %>%
%   phylobase::subset(mrca = basids$seq) %>%
%   as("phylo")
% 
% basids2 <- treetax %>% filter(seq %in% basidtree$tip.label)
% 
% ggtree(groupOTU(basidtree, split(basids2$seq, basids2$Order)), layout = "circular") +
%   aes(color = group) +
%   scale_color_brewer(type = "qual", palette = "Paired", limits = unique(na.omit(basids2$Order))) +
%   theme(legend.position = NULL)
% @
% \end{frame}

\begin{frame}{Correlogram}

<<correlogram, fig.width = 12>>=
bind_rows(as.data.frame(bc_corr_long$mantel.res) %>%
            mutate(tdist = "same", seq_run = "long"),
          as.data.frame(bc_corr_long2$mantel.res) %>%
            mutate(tdist = "diff", seq_run = "long"),
          as.data.frame(bc_corr_short$mantel.res) %>%
            mutate(tdist = "same", seq_run = "short"),
          as.data.frame(bc_corr_short2$mantel.res) %>%
            mutate(tdist = "diff", seq_run = "short")) %>%
  filter(complete.cases(.)) %>%
  mutate(Significant = `Pr(corrected)` < 0.05) %>%
  ggplot(aes(class.index, Mantel.cor, group = tdist, color = tdist, shape = Significant)) +
  geom_point(size = 3) +
  geom_line(alpha = 0.3, size = 1.5) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  # geom_line(aes(y = exp(predict(bc_fit)))) +
  scale_y_continuous(limits = c(-0.05, NA), name = "Spatial Distance correlation") +
  scale_color_discrete(name = "Year") +
  xlab("distance (m)") +
  facet_wrap("seq_run") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1))
@
\end{frame}

\begin{frame}{Acknowledgements}
\begin{columns}[t]
  \begin{column}{0.5\linewidth}
    \textbf{Uppsala University}\\
    Anneli Svanholm\\
    Bobby Sulistyo\\
    Ylva Strid\\
    Kerri Kluting
  \end{column}
  \begin{column}{0.5\linewidth}

    \textbf{University of Parakou}\\
    Roel Houdanon\\
    Evans Codjia\\
    Souleymane Boni\\
    Hyppolite Aignon\\
    Sylvestre Badou\\
    Gerard Laourou\\
    Aziz Boukary\\
    Olivia Fayedi
  \end{column}
\end{columns}
\includegraphics[width=0.5\linewidth]{formas}
\end{frame}
